Gravitational waves from colliding black holes

### 2017-05-05 Report/Draft/draft.tex
Latest

	\HRule \\[0.4cm]
	{ \huge \bfseries Gravitational waves from colliding black holes}\\[0.4cm] % Title of your document
	\HRule \\[1.5cm]

@@

\begin{minipage}{0.4\textwidth}
	\begin{flushleft} \large
		\emph{Author:}\\
		Christian \textsc{Pedersen} % Your name
	\end{flushleft}
\end{minipage}
~
\begin{minipage}{0.4\textwidth}
	\begin{flushright} \large
		\emph{Supervisor:} \\
		Prof. Stephen \textsc{Fairhurst} % Supervisor's Name
	\end{flushright}
\end{minipage}\\[1cm]
	% If you don't want a supervisor, uncomment the two lines below and remove the section above
	%\Large \emph{Author:}\\
	%John \textsc{Smith}\\[3cm] % Your name
	Student number: 1265898\\
	Primary Assessor: Prof. Carole \textsc{Tucker}\\
	Secondary Assessor: Dr. Clarence \textsc{Matthai}\\
	%----------------------------------------------------------------------------------------
	%	DATE SECTION
	%----------------------------------------------------------------------------------------
	{\large \today}\\[1cm] % Date, change the \today to a set date if you want to be precise

	%----------------------------------------------------------------------------------------
	%	LOGO SECTION
	%----------------------------------------------------------------------------------------

	%\includegraphics{Logo}\\[1cm] % Include a department/university logo - this will require the graphicx package

@@

@@

Given that so many parameters, many of them extrinsic, describe only two sets of timeseries data (one from each detector), parameter estimation is a challenging task and the parameter space is both enormous and wrought with degenerecies. Two of the more thorougly researched degenerecies are those between total mass, distance and inclination, as all three parameters scale the amplitude of the signal and between mass and spin\cite{spindegen}, which is a degenerecy that arises out of post-Newtonian (PN) theory.

The qualitative idea behind current methods of parameter estimation is that the GW signal is processed and extracted from the raw strain data, and then matched against an array of simulated waveforms to find which waveform most closely resembles the detected signal. The two technical challenges here are quantifying how well a template represents the observed signal, and how to efficiently sample the parameter space for new templates to test against the data. Within LIGO a standard method for quantifying the similarity between two signals $h_1(t)$ and $h_2(t)$ is given by the noise weighted inner product, also referred to as the 'match' between signals, and is defined as
\begin{equation}
\langle h_1(t) \vert h_2(t) \rangle=2\int_{0}^{\infty}\frac{\tilde{h_1^*}(f)\tilde{h_2}(f)+\tilde{h_1}(f)\tilde{h_2^*}(f)}{S_n(f)}df 
\end{equation}
where $\tilde{h}(f)$ is the Fourier transform of the signal $h(t)$, and $S_n(f)$ is the noise curve of the detector\cite{oldest_pe}. It is also convenient here to define the term \textit{mismatch}, which is simply $1-m$ where $m$ is the match. A variety of methods have been employed to sample the parameter space, including nested sampling\cite{pe2} and analysis using Gaussian wavelets\cite{props}, and in this work we use a framework of Bayesian inference and Markov-Chain Monte Carlo methods\cite{inj}\cite{pe}\cite{pe4}. This process results in a set of posterior distribution functions (PDFs) for each parameter. Using Bayes' Theorem, the posterior is given by
\begin{equation}
p(\vec{\theta})|d)=\frac{p(\vec{\theta})p(d|\vec{\theta})}{P(d)}
\end{equation}
where $\vec{\theta}$ is an $n$ dimensional vector in our parameter space of $n$ parameters\cite{pe2}. The first term in the numerator, $p(\vec{\theta})$ is known as the \textit{prior} distribution, and quantifies knowledge we already have about the system that should influence our estimation of its' parameters. In practice, in inference of BBHs, the priors are almost always uniform or isotropic distributions. The denominator, $P(d)$, is essentially a normalisation factor ensuring that the posterior distribution integrats to unity. The crucial term is the \textit{likelihood} for a given set of parameters given the data, which is determined by
\begin{equation}
p(d|\vec{\theta})\propto \exp\bigg(-\frac{1}{2}\sum_{k=1,2}\Big \langle h^M_k(\vec{\theta})-d_k \big\vert h^M_k(\vec{\theta})-d_k \Big \rangle\bigg)
\end{equation}
where the inner product is that defined in (13), $d_k$ is the observed strain in the detector, and $h_k^M(t;\vec{\theta})$ is the simulated detector response for a given set of parameters $\vec{\theta}$ given by (9). The SNR is given by

@@

The particular focus of this paper is on the challenges of inferring $\chi_p$ in precessing systems. This parameter is of particular importance due to its astrophysical implications, and as of yet there has been no comprehensive study into the effects of precession and it's amplitude modulation of the signal on the process of parameter estimation. The formation methods of compact binary systems of stellar mass black holes are currently unknown, and a variety of models have been proposed\cite{modal}. Of particular interest is whether the two black holes formed from a common accreting system, or whether the binary was formed by dynamical capture. The former model would imply that the spins on the black holes would generally tend to be aligned with one another and with the orbital angular momentum, however in the dynamical capture model we expect the spins to be more or less uniformly distributed. Given the large number of expected detections over the coming years, a thorough understanding of the detector's response to precessing waveforms and an accurate estimation of our ability to recover $\chi_p$ reliably will be crucial to answering these questions, and maximising the scientific yield from this remarkable technology.
\section{Intrinsic degeneracies}
\subsection{Precessing waveforms}
We first consider in more detail the effects of precession on a GW waveform. This is important as understanding how precessive waveforms behave is key to understanding how the detector will respond to these signals, and understanding the detector response is the fundamental idea of parameter estimation. A closer look at equation (16), which essentially quantifies the amount of precession in a given binary, shows that it takes values in the range $[0,1]$, with the maximum reached when the spin on the larger BH is fully in-plane. Even for a maximally precessing signal however, many possible configurations of polarisation and inclination are possible which will affect the way the amplitude of the signal is modulated.

@@

Fig. 10 shows how the matches between precessing and non-precessing waveforms change with inclination for four different mass ratios. The dark regions of the match is where the signals have significant mismatch, and the waveforms should be well distinguished, and hte light regions are where the waveforms are effectively degenerate. It is apparent that the degeneracy is far stronger for close to equal mass binaries, where only maximally precessing systems at edge-on inclinations will have noticeably different waveforms. Here we plot the 0.97 match contour as it is generally expected that the advanced LIGO detectors should be able to distinguish between waveforms with a match of 0.97. For all but the most extreme mass ratios, a combination of significant in-plane spin ($\geq0.6$) and/or strong inclination is necessary to have any significant deviation for systems with high precession. It is important to remember here that as we change the inclination, the overall strength of the signal will drop as the amplitude of the gravitational wave is lower (with the peak at $\iota=0$), and the matching process does not account for this change in signal strength.

@@

Lastly, in Fig. 12 we explore the impact of polarisation and phase on the parameter space, using the same technique as in Fig. 11. We select a high mass ratio binary here ($m1=55, m2=15$) to highlight any prominent features of the parameter space. A full range of inclinations and phases are shown for four polarisations: plus, mixed but plus dominated, mixed but cross domianted, and cross polarisations. The region of extremely high sensitivity at $\iota=\pi/2$ is a result of the fact that there is no cross-polarised signal for an edge-on binary. The parameter space appears to be highly structured, and dependent on both phase and inclination. Overall the parameter space appears a lot smoother for plus polarisations, and variations across it increase as we shift to a cross polarised signal. As such we can expect to resolve spin parameters slightly better in cross polarised waveforms. The figures also show that phase does have an effect on spin sensitivity, with small shifts in phases changing the spin magnitudes by up to $\approx0.3$.

@@

\section{Conclusions}
Overall the results indicate that the current inference methods consistently underestimate values of $\chi_p$, with all but 5 out of the 20 inference runs excluding the injected value of $\chi_p$. The match results and waveform plots demonstrate that the phase of the GW signal is an important parameter in precessing systems unlike in non-precessing systems, leading to an extra degree of freedom in precessing systems. The matching process was used to highlight the way in which mass ratios, polarisations, inclinations and phases affect precessing waveforms, particularly our ability to distinguish bewteen precessing and non-precessing signals. Degenerate areas of the parameter space were also identified using this technique, and it was shown that near for equal mass ratios, precessing and non-precessing waveforms are effectively entirely degenerate. It was also shown that other parameters such as source location, polarisation and inclination affect the way that the modulation effect manifests itself in a specific signal. This leads to a key observation that the strength of signal modulation in a GW waveform is not necessarily proportional to the amount of precession in the binary itself.

The inference pipeline used to perform parameter estimation was then described, and a series of injections with high mass ratio and high $\chi_p$ were injected into detector noise. The standard inference process was performed, and results were compared for a case of $\chi_p$ being recovered comparably well, and $\chi_p$ being excluded. Neither sets of posteriors were fully accurate, but the case of $\chi_p$ being excluded actually provided more accurate results on the whole. Inference results were presented for a signal which appeared to exhibit high precession, however the MAP values on this signal still exluded the injected value of $\chi_p$ while still showing a strong match (0.96) with the injected signal. Posteriors of derived spin parameters were shown for a full range of inclinations for signals at 200Mpc and 800Mpc, and their results were compared with each other and with the predictions made by the matches. No clear correlations between any of these were observed, indicating that the relationship between matches and inference results is more complicated than previously thought. Finally a set of inference runs includingthe Virgo detector at a range of inclinations were presented and compared with previous results without Virgo. These results suggested that Virgo will not significantly improve the accuracy on $\chi_p$ due to the fact that the morphology of the signal and the amplitude modulation is the same in the Virgo detector as in H1 and L1. Although the specific set of parameters used in these injections greatly limits the generality of this result.

@@

There is significantly more work to be done in this area, and this comes nowhere close to a conclusive study. Due to time constraints, this work only explored a very small sliver of the overall parameter space. However, while it is certainly possible that the results presented here are specific to this region, given that the trend of exluding $\chi_p$ is so strong in these results, it is unlikely to be entirely unique. During the project, individual software injections were performed for a wider range of parameters than those shown here and the general trend of $\chi_p$ being excluded was observed there as well, however this is now anecdotal evidence and much closer analysis is needed, both in the form of larger scope studies of the parameter space and more targeted studies to understand better the causes of incorrect parameter inference. As a final point, many of the posteriors of $\chi_p$ shown in this work are extremely similar to those shown in the recent detection papers\cite{obs}\cite{obs2} despite the fact that in many cases the injected parameters were excluded by the inference process. If there is indeed a bias here towards certain spin orientations or magnitudes, it is crucial that this is identified in order ensure that astrophysical conclusions drawn from current and future LIGO observations are well grounded.
\section{Acknowledgements}
I would firstly like to thank Steve Fairhurst for his patient and warm supervision. Our meetings were always informative and enjoyable and I have learned a lot from him over the past year. I would also like to give a huge thanks to Collin Capano, without whom this report would have been significantly shorter, for being extremely generous with his time and helping me navigate the labyrinth of LIGO codes. Finally I would like to thank Callum Booth for being my partner in crime over the last 9 months.
\section{Reflective Statement}
This project has certainly taught me the value of time, and what a scarce resource it is. These kinds of things seem to suck up an endless amount of time for very little apparent progress, and knowing when to call it and do something else for a few hours is something I'm not so great at. Something I need to learn in the future is to figure out a way to switch off and reset once in a while. Overall I did have a great time working on it. The LIGO collaboration is truly awesome and this was a really great opportunity to be a part of it. I certainly learnt a lot of useful technical skills, especially learning to use GitHub from the command line, and my programming improved dramatically over the course of the project.

>>>>

	\HRule \\[0.2cm]
	{ \huge \bfseries Gravitational waves from colliding black holes}\\[0.4cm] % Title of your document
	\HRule \\[1.cm]

@@

\begin{minipage}{0.4\textwidth}
	\begin{flushleft} \large
		\emph{Author:}\\
		Christian \textsc{Pedersen} % Your name
	\end{flushleft}
\end{minipage}
~
\begin{minipage}{0.4\textwidth}
	\begin{flushright} \large
		\emph{Supervisor:} \\
		Prof. Stephen \textsc{Fairhurst} % Supervisor's Name
	\end{flushright}
\end{minipage}\\[1cm]

	%\begin{tabular}{l*{2}{c}r}
	%	Student number: & 1265898 \\
%		Degree Programme:&  MPhys Physics with Astronomy \\
%		Primary Assessor:& Prof. Carole \textsc{Tucker} \\
%		Secondary Assessor:& Dr. Clarence \textsc{Matthai} \\
%	\end{tabular}

\begin{align*}
\text{\emph{Student Number}}&: \text{1265898} \\
\text{\emph{Degree Programme}}&: \text{MPhys Physics with Astronomy} \\
\text{\emph{Primary Assessor}}&: \text{Prof. Carole \textsc{Tucker}} \\
\text{\emph{Secondary Assessor}}&: \text{Dr. Clarence \textsc{Matthai}} \\
\end{align*}
%\begin{tabular}{2}
%Student number: &1265898\\
%Degree Programme:& MPhys Physics with Astronomy\\
%Primary Assessor:& Prof. Carole \textsc{Tucker}\\
%
%\end{tabular}

@@

	%----------------------------------------------------------------------------------------
	{\large \today}\\[1cm] % Date, change the \today to a set date if you want to be precise

	%----------------------------------------------------------------------------------------
\begin{flushleft}
\textbf{Declaration:}I have read and understand Appendix 2 in the Student Handbook: “Some advice on the avoidance of plagiarism”. I hereby declare that the attached 
dissertation is exclusively my own work, that no part of the work has previously been 
submitted for assessment, and that I have not knowingly allowed it to be copied by another person.
\end{flushleft}

@@

Given that so many parameters, many of them extrinsic, describe only two sets of timeseries data (one from each detector), parameter estimation is a challenging task and the parameter space is both enormous and wrought with degenerecies. Two of the more thoroughly researched degenerecies are those between total mass, distance and inclination, as all three parameters scale the amplitude of the signal and between mass and spin\cite{spindegen}, which is a degenerecy that arises out of post-Newtonian (PN) theory.

The qualitative idea behind current methods of parameter estimation is that the GW signal is processed and extracted from the raw strain data, and then matched against an array of simulated waveforms to find which waveform most closely resembles the detected signal. The two technical challenges here are quantifying how well a template represents the observed signal, and how to efficiently sample the parameter space for new templates to test against the data. Within LIGO a standard method for quantifying the similarity between two signals $h_1(t)$ and $h_2(t)$ is given by the noise weighted inner product, also referred to as the 'match' between signals, and is defined as
\begin{equation}
\langle h_1(t) \vert h_2(t) \rangle=2\int_{0}^{\infty}\frac{\tilde{h_1^*}(f)\tilde{h_2}(f)+\tilde{h_1}(f)\tilde{h_2^*}(f)}{S_n(f)}df 
\end{equation}
where $\tilde{h}(f)$ is the Fourier transform of the signal $h(t)$, and $S_n(f)$ is the noise curve of the detector\cite{oldest_pe}. It is also convenient here to define the term \textit{mismatch}, which is simply $1-m$ where $m$ is the match. A variety of methods have been employed to sample the parameter space, including nested sampling\cite{pe2} and analysis using Gaussian wavelets\cite{props}, and in this work we use a framework of Bayesian inference and Markov-Chain Monte Carlo methods\cite{inj}\cite{pe}\cite{pe4}. This process results in a set of posterior distribution functions (PDFs) for each parameter. Using Bayes' Theorem, the posterior is given by
\begin{equation}
p(\vec{\theta})|d)=\frac{p(\vec{\theta})p(d|\vec{\theta})}{P(d)}
\end{equation}
where $\vec{\theta}$ is an $n$ dimensional vector in our parameter space of $n$ parameters\cite{pe2}. The first term in the numerator, $p(\vec{\theta})$ is known as the \textit{prior} distribution, and quantifies knowledge we already have about the system that should influence our estimation of its' parameters. In practice, in inference of BBHs, the priors are almost always uniform or isotropic distributions. The denominator, $P(d)$, is essentially a normalisation factor ensuring that the posterior distribution integrates to unity. The crucial term is the \textit{likelihood} for a given set of parameters given the data, which is determined by
\begin{equation}
p(d|\vec{\theta})\propto \exp\bigg(-\frac{1}{2}\sum_{k=1,2}\Big \langle h^M_k(\vec{\theta})-d_k \big\vert h^M_k(\vec{\theta})-d_k \Big \rangle\bigg)
\end{equation}
where the inner product is that defined in (13), $d_k$ is the observed strain in the detector, and $h_k^M(t;\vec{\theta})$ is the simulated detector response for a given set of parameters $\vec{\theta}$ given by (9). The SNR is given by

@@

The particular focus of this paper is on the challenges of inferring $\chi_p$ in precessing systems. This parameter is of particular importance due to its astrophysical implications, and as of yet there has been no comprehensive study into the effects of precession and its amplitude modulation of the signal on the process of parameter estimation. The formation methods of compact binary systems of stellar mass black holes are currently unknown, and a variety of models have been proposed\cite{modal}. Of particular interest is whether the two black holes formed from a common accreting system, or whether the binary was formed by dynamical capture. The former model would imply that the spins on the black holes would generally tend to be aligned with one another and with the orbital angular momentum, however in the dynamical capture model we expect the spins to be more or less uniformly distributed. Given the large number of expected detections over the coming years, a thorough understanding of the detector's response to precessing waveforms and an accurate estimation of our ability to recover $\chi_p$ reliably will be crucial to answering these questions, and maximising the scientific yield from this remarkable technology.
\section{Intrinsic degeneracies}
\subsection{Precessing waveforms}
We first consider in more detail the effects of precession on a GW waveform. This is important as understanding how precessive waveforms behave is key to understanding how the detector will respond to these signals, and understanding the detector response is the fundamental idea of parameter estimation. A closer look at equation (16), which essentially quantifies the amount of precession in a given binary, shows that it takes values in the range $[0,1]$, with the maximum reached when the spin on the larger BH is fully in-plane. Even for a maximally precessing signal however, many possible configurations of polarisation and inclination are possible which will affect the way the amplitude of the signal is modulated.

@@

Fig. 10 shows how the matches between precessing and non-precessing waveforms change with inclination for four different mass ratios. The dark regions of the match are where the signals have significant mismatch, and the waveforms should be well distinguished, and the light regions are where the waveforms are effectively degenerate. It is apparent that the degeneracy is far stronger for close to equal mass binaries, where only maximally precessing systems at edge-on inclinations will have noticeably different waveforms. Here we plot the 0.97 match contour as it is generally expected that the advanced LIGO detectors should be able to distinguish between waveforms with a match of 0.97. For all but the most extreme mass ratios, a combination of significant in-plane spin ($\geq0.6$) and/or strong inclination is necessary to have any significant deviation for systems with high precession. It is important to remember here that as we change the inclination, the overall strength of the signal will drop as the amplitude of the gravitational wave is lower (with the peak at $\iota=0$), and the matching process does not account for this change in signal strength.

@@

Lastly, in Fig. 12 we explore the impact of polarisation and phase on the parameter space, using the same technique as in Fig. 11. We select a high mass ratio binary here ($m1=55, m2=15$) to highlight any prominent features of the parameter space. A full range of inclinations and phases are shown for four polarisations: plus, mixed but plus dominated, mixed but cross dominated, and cross polarisations. The region of extremely high sensitivity at $\iota=\pi/2$ is a result of the fact that there is no cross-polarised signal for an edge-on binary. The parameter space appears to be highly structured, and dependent on both phase and inclination. Overall the parameter space appears a lot smoother for plus polarisations, and variations across it increase as we shift to a cross polarised signal. As such we can expect to resolve spin parameters slightly better in cross polarised waveforms. The figures also show that phase does have an effect on spin sensitivity, with small shifts in phases changing the spin magnitudes by up to $\approx0.3$.

@@

\section{Conclusions}
Overall the results indicate that the current inference methods consistently underestimate values of $\chi_p$, with all but 5 out of the 20 inference runs excluding the injected value of $\chi_p$. The match results and waveform plots demonstrate that the phase of the GW signal is an important parameter in precessing systems unlike in non-precessing systems, leading to an extra degree of freedom in precessing systems. The matching process was used to highlight the way in which mass ratios, polarisations, inclinations and phases affect precessing waveforms, particularly our ability to distinguish bewteen precessing and non-precessing signals. Degenerate areas of the parameter space were also identified using this technique, and it was shown that for near equal mass ratios, precessing and non-precessing waveforms are effectively entirely degenerate. It was also shown that other parameters such as source location, polarisation and inclination affect the way that the modulation effect manifests itself in a specific signal. This leads to a key observation that the strength of signal modulation in a GW waveform is not necessarily proportional to the amount of precession in the binary itself.

The inference pipeline used to perform parameter estimation was then described, and a series of injections with high mass ratio and high $\chi_p$ were injected into detector noise. The standard inference process was performed, and results were compared for a case of $\chi_p$ being recovered comparably well, and $\chi_p$ being excluded. Neither sets of posteriors were fully accurate, but the case of $\chi_p$ being excluded actually provided more accurate results on the whole. Inference results were presented for a signal which appeared to exhibit high precession, however the MAP values on this signal still excluded the injected value of $\chi_p$ while still showing a strong match (0.96) with the injected signal. Posteriors of derived spin parameters were shown for a full range of inclinations for signals at 200Mpc and 800Mpc, and their results were compared with each other and with the predictions made by the matches. No clear correlations between any of these were observed, indicating that the relationship between matches and inference results is more complicated than previously thought. Finally a set of inference runs including the Virgo detector at a range of inclinations were presented and compared with previous results without Virgo. These results suggested that Virgo will not significantly improve the accuracy on $\chi_p$ due to the fact that the morphology of the signal and the amplitude modulation is the same in the Virgo detector as in H1 and L1. Although the specific set of parameters used in these injections greatly limits the generality of this result.

@@

There is significantly more work to be done in this area, and this comes nowhere close to a conclusive study. Due to time constraints, this work only explored a very small sliver of the overall parameter space. However, while it is certainly possible that the results presented here are specific to this region, given that the trend of excluding $\chi_p$ is so strong in these results, it is unlikely to be entirely unique. During the project, individual software injections were performed for a wider range of parameters than those shown here and the general trend of $\chi_p$ being excluded was observed there as well, however this is now anecdotal evidence and much closer analysis is needed, both in the form of larger scope studies of the parameter space and more targeted studies to understand better the causes of incorrect parameter inference. As a final point, many of the posteriors of $\chi_p$ shown in this work are extremely similar to those shown in the recent detection papers\cite{obs}\cite{obs2} despite the fact that in many cases the injected parameters were excluded by the inference process. If there is indeed a bias here towards certain spin orientations or magnitudes, it is crucial that this is identified in order ensure that astrophysical conclusions drawn from current and future LIGO observations are well grounded.
\section{Acknowledgements}
I would firstly like to thank Steve Fairhurst for his patient and warm supervision. Our meetings were always informative and enjoyable and I have learned a lot from him over the past year. I would also like to give a huge thanks to Collin Capano, without whom this report would have been significantly shorter, for being extremely generous with his time and helping me navigate the labyrinth of LIGO codes. Finally I would like to thank Callum Booth for being my partner in crime over the last 9 months.
\section{Reflective Statement}
This project has certainly taught me the value of time, and what a scarce resource it is. These kinds of things seem to suck up an endless amount of time for very little apparent progress, and knowing when to call it and do something else for a few hours is something I'm not so great at. Something I need to learn in the future is to figure out a way to switch off and reset once in a while. Overall I did have a great time working on it. The LIGO collaboration is truly awesome and this was a really great opportunity to be a part of it. I certainly learnt a lot of useful technical skills, especially learning to use GitHub from the command line, and my programming improved dramatically over the course of the project.
